Syntax Is All You Need: A Universal-Language Approach to Mutant Generation

### 2023-09-25 paper/main.tex
Alex pass one

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handle
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).

@@

\subsection{Incremental Mutation}

One obvious concern with any-language source-based mutation is that every time the
program changes, the cost of invalid mutants must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.

The {\tt git merge-file} utility takes a base file and two modified
versions of that file, and produces an automatic merge, using the
usual git merge resolution algorithm.    Because a mutant is a (very
small) source change, this means that in practice updating a mutant to
reflect even large changes to a source file is cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  To the extent development is incremental,
therefore, mutant generation can also be incremental.  Our
implementation allows the use of {\tt git merge-file} to control the
modification of most mutants, and uses our mechanism for
selective mutant generation based on coverage to force mutation where
there is a conflict in a ``mutant patch.''

@@

We speculate that our somewhat lower scores and possibly higher
equivalence rates derive from the simple fact that we generate more
mutants than most tools.  The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using formal
verification, has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.

The price paid for this
expansiveness in allowing ways to detect weaknessess in testing is a
larger number of mutants to execute, and a somewhat larger number of
equivalent mutants.  This cost, we observe, falls especially on Comby in some
cases, because Comby is better at identifying places to inject
control flow changes that produce valid code, but do yield equivalent mutants in
many cases.

However, for our approach, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into (at least) two sets, one
set of core operators similar to (but larger than) those provided by
LittleDarwin or Cargo-Mutants, and a set of ``aggressive'' operators
that includes our full current set of operators, a very simple matter due to
our approach to mutant generation.

@@

Finally, the practical use of mutation testing at Google argues that
support for a variety of languages is critical.  Google has used its
substantial resources to provide mutant generation for C++, Java,
Python, Javascript, Go, Typescript, and Common
Lisp~\cite{PetrovicMutationGoogle}, all of which we also support.  The
operators Google uses for these
languages (based on the original Mothra operators~\cite{offutt1996experimental}) are a subset of those provided by our implementation, with the caveat
that block removal is only supported when using Comby;
regular expression mutation is limited to single-line statement
deletion.  For Lisp-like languages, the difficulty of identifying statements vs. value-returning
function calls makes use of block deletion somewhat impractical.  The majority of the Google Mothra operators
can be implemented as universal rules for all languages; only specializing
logical operators and implementing block/statement deletion even requires
language identification.  Unlike Google's mutation infrastructure, our
tool is open source and easily extensible.

The approach reported in this paper was briefly presented in part in a
tool paper (citation redacted for blinding purposes) that introduced
the idea of using regular expressions to generate mutants for multiple
languages, and provided an evaluation against PIT for a single toy
Java program.  The present paper provides the full idea and context,
provides empirical data for universal mutant generation on real world
programs over four languages,
and introduces the idea of using parser parser combinators to go
beyond regular expression-based generation.

>>>>

      \item A comparison of an implementation of mutant generation
       based on this approach with existing tools for four important
       programming languages.  The single-language tools range from
       approximately 3,800 LOC to nearly 60,000 LOC, and, we emphasize, each handle
       one programming language.  Our tool, which provides comparable
       core mutation testing functionality, supports over a dozen
       languages using only about 2,200 LOC of Python and less than 500
       lines for rules defining mutation operators over a dozen
       specific languages (and which can effectively mutate an
       essentially unlimited number of other languages).  Our
       approach, we emphasize, enables a smaller tool than any other tool
       we examined (by over 1000 LOC) to support a rich set of
       operators (the full set used by Google's internal mutation
       tools, plus a large number of additional operators) for all languages covered by other tools, plus a set
       of additional languages.

@@

\subsection{Supported Languages and Operators}

Google has used its substantial resources to provide mutant generation for C++, Java,
Python, Javascript, Go, Typescript, and Common
Lisp~\cite{PetrovicMutationGoogle}, all of which we also support.  The
operators Google uses for these
languages (based on the original Mothra operators~\cite{offutt1996experimental}) are a subset of those provided by our implementation, with the caveat
that block removal is only supported when using Comby;
regular expression mutation is limited to single-line statement
deletion.  For Lisp-like languages, the difficulty of identifying statements vs. value-returning
function calls makes use of block deletion somewhat impractical.  The majority of the Google Mothra operators
can be implemented as universal rules for all languages; only specializing
logical operators and implementing block/statement deletion even requires
language identification.  In addition to these core operators, we
provide a number of additional operators for every language offered,
and as shown below these produce a richer (in terms of operators used) set of mutants than most
tools currently used.

There is generally a one-to-one
correspondence between Comby and Regexp rules in our implementation.
The number of Comby rules (which is a small multiple of the number of
``operators'' being defined ) for languages discussed below, at each hierarchical level in our
implementation, is:  75 (universal), 33 (Python), 16 (C-like), 14
(C++), 7 (Java), and 2 (Rust).

The primary limitation of our tool is that it does not offer operators
that require semantic knowledge of the program beyond the
identification of roles offered by Comby.
The only operator from the mutation testing literature that we are aware of that
cannot be supported by our approach as a result, to our knowledge, is variable
replacement.  However, we are aware of no widely-used tool or even recent
study of mutation testing that makes use of this type of operator.

One obvious concern with any-language source-based mutation is that every time the
program changes, the cost of invalid mutants must be paid again.
However, the locality of mutants and source changes in fact means this
is seldom required.   Because a mutant is a (very
small) source change,in practice updating a mutant to
reflect even large changes to a source file is nearly cost-free, and if the
original mutant was valid, the new mutant will usually also be valid.
New mutant generation is only required for new lines of code and
modified lines of code.  Our
implementation allows the use of the {\tt git merge-file} algorithm to control the
modification of most mutants, and uses our mechanism for
selective mutant generation to force mutation where
there is a conflict in a ``mutant patch.''

@@

We speculate that our somewhat lower scores and possibly higher
equivalence rates derive from the simple fact that \emph{we generate
substantially more (non-equivalent and overall)}
mutants than most tools.  In fact, for every single language studied,
our tool generated the largest number of mutants, and without a
notably higher or unacceptable equivalence rate, or a notably lower
mutation score.  Our mean number of generated valid, non-equivalent
mutants (estimating based on our sample) was more than a factor of
two larger than the nearest competing tool for all languages except
Python, and was still more than 300 mutants larger than the nearest
competing tool for Python.

The ease of defining mutation operators,
plus our interest in generating large numbers of mutants and
prioritizing the most interesting, in order to make mutation testing useful for
projects with extremely high quality test suites, or even using formal
verification, has encouraged us to take a ``let a thousand flowers bloom''
approach to choosing mutation operators.  For example, our approach
provides many more control-flow mutations than most tools, due to our
observation that test suites sometimes fail to check for loop
execution counts other than one and zero.  Few other tools introduce
{\tt break} and {\tt continue} statements into programs, or transform
one into the other.  These mutants are often equivalent (in fact, they
give rise to a large number of our equivalent mutants reported above), but \emph{also
sometimes expose subtle weaknesses in a high quality test suite}, in our
experience.  In general, we take our philosophy of mutation from the
proposal by Gopinath et al.~\cite{Limits,gopinath2017mutation} that
expanding the set of mutation operations and classes of such
operations, in the long run, is more important than using only a small
set of ``good'' operators, in that the end-goal of mutation testing is
to identify weaknesses in test suites, not simply to estimate a single
score.   The price paid for this
expansiveness in allowing ways to detect weaknessess in testing is a
larger number of mutants to execute, and a slightly larger number of
equivalent mutants.

%This cost, we observe, falls especially on Comby in some
%cases, because Comby is better at identifying places to inject
%control flow changes that produce valid code, but do yield equivalent mutants in
%many cases.

However, this is not a
fundamental limitation or forced choice.  We propose as future work to
split the operators for each level of the hierarchy into (at least) two sets, one
set of core operators similar to (but larger than) those provided by
LittleDarwin or Cargo-Mutants, and a set of ``aggressive'' operators
that includes our full current set of operators, a very simple matter due to
our approach to mutant generation.

@@

Finally, the practical use of mutation testing at Google argues that
support for a variety of languages is critical.  As noted above, we
cover the set of languages and operators used in Google's internal
mutation testing tools~\cite{PetrovicMutationGoogle}, but in an open
source, easily extensible tool offering a large number of additional
languages and operators.

% The approach reported in this paper was briefly presented in part in a
% tool paper (citation redacted for blinding purposes) that introduced
% the idea of using regular expressions to generate mutants for multiple
% languages, and provided an evaluation against PIT for a single toy
% Java program.  The present paper provides the full idea and context,
% provides empirical data for universal mutant generation on real world
% programs over four languages,
% and introduces the idea of using parser parser combinators to go
% beyond regular expression-based generation.
